Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-11-09 text/ms.tex
Add 'downward events' in discussion summary
Edits to ms

The majority of our heavy-tailed estimates were robust to alternative population models, observation error, and choice or priors. Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates without density dependence, or modelled the population dynamics as Ricker-logistic (Fig.~\ref{fig:alt}). However, setting observation error standard deviation to $0.2$ increased the median estimate of $\nu$ from $<10$ to $\ge 10$ in $\baseNuTenObsTenSwitch$ of $\baseNuTen$ populations, although the majority of $\nu$ estimates remained qualitatively similar (Fig.~\ref{fig:alt}). The strength of the prior on $\nu$ had little influence on estimates of black-swan dynamics (Fig.~\ref{fig:alt-priors}). Our simulation testing shows that, if anything, our models underpredict the true magnitude and probability of heavy tailed events --- especially given the length of the time series in the GPDD (Figs~\ref{fig:sim-nu}, \ref{fig:sim-prob}).

Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Figs~\ref{fig:correlates}, \ref{fig:posteriors}a). Longer time-series length was the strongest covariate of observing black-swan dynamics. For example, we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails in a population with 60 time steps than in one with 30 time steps. However, the absolute change in probability with increased time series length was small ($\pHeavyNSixty$ vs.\ $\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).

We examined all time series with published explanations of why the black-swan events occurred (Tables~\ref{tab:causes-supp} and \ref{tab:sparks}). The majority (\percBSDown \%) of documented events were downward black swans and involved a series of unfortunate events (Table~\ref{tab:causes-supp}). For example, a shortage of nest sites for the shag long-necked bird (\textit{Phalacrocorax aristotelis}) in the United Kingdom reduced population productivity and made the population more vulnerable when a red tide hit in 1968 \citep{potts1980}. The population experienced a black-swan downswing, but the new availability of nest sites resulted in a rapid population upswing \citep{potts1980}. As another example, an interaction between environmental- and predation-mediated population cycles are thought to have caused a downward black-swan event for a water vole (\textit{Arvicola terrestris}) population \citep{saucy1994}. Other black swans were the result of a sequence of extreme climate events on their own. For example, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\textit{Ardea cinerea}) abundance in the United Kingdom \citep{stafford1971}. Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted \citep{stafford1971}. \textit{(TODO: substitute a parasite example.)}

@@

We found strong evidence for black swans (heavy-tailed process noise) in 3--5\% of ecological time series. Black swans are detected more frequently in longer time series and in populations with a smaller magnitude of process error. However, black swans were not associated with density dependence, population growth rate, or lifespan. In verified cases, black swans were often a result of the interaction between elements of extreme climate, predators and parasite cycles, and strong changes in human pressures. Our empirical results, sensitivity analyses, and simulation tests suggest that estimating the tail shape of process noise is a viable method of detecting black-swan population dynamics and if anything will underestimate the probability of black swans. The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century \citep{easterling2000,ipcc2012}.

@@

There are a number of caveats when considering the generality of our results. The GPDD data represent a taxonomically and geographically biased sample of populations --- the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom. Although we expect that we would find qualitatively similar evidence for black swans in many other large taxonomic or geographic samples of populations, the common forces driving those black swans likely differ (e.g.~severe winters in Table~\ref{tab:sparks}). In addition to a possibly biased sample of populations, some black-swan detections could just be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes. A final caveat is that the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and this likely influences the detection of heavy tails. For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black swans. Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.

>>>>

The majority of our heavy-tailed estimates were robust to alternative population models, observation error, and choice of priors. Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates without density dependence, or modelled the population dynamics as Ricker-logistic (Fig.~\ref{fig:alt}). However, setting observation error standard deviation to $0.2$ increased the median estimate of $\nu$ from $<10$ to $\ge 10$ in $\baseNuTenObsTenSwitch$ of $\baseNuTen$ populations, although the majority of $\nu$ estimates remained qualitatively similar (Fig.~\ref{fig:alt}). The strength of the prior on $\nu$ had little influence on estimates of black-swan dynamics (Fig.~\ref{fig:alt-priors}). Our simulation testing shows that, if anything, our models underpredict the true magnitude and probability of heavy tailed events --- especially given the length of the time series in the GPDD (Figs~\ref{fig:sim-nu}, \ref{fig:sim-prob}).

Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Figs~\ref{fig:correlates}, \ref{fig:posteriors}a). Longer time-series length was the strongest covariate of observing black-swan dynamics. For example, we were about \pIncHeavyNThirtyNSixty~times as likely to observe heavy tails in a population with 60 time steps compared to one with 30 time steps. However, the absolute change in probability with increased time series length was small ($\pHeavyNSixty$ vs.\ $\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).

We examined all time series with published explanations of why the black-swan events occurred (Tables~\ref{tab:sparks} and \ref{tab:causes-supp}). The majority (\percBSDown \%) of documented events were downward black swans and involved a series of unfortunate events (Table~\ref{tab:causes-supp}). For example, a shortage of nest sites for the shag long-necked bird (\textit{Phalacrocorax aristotelis}) in the United Kingdom reduced population productivity and made the population more vulnerable when a red tide hit in 1968 \citep{potts1980}. The population experienced a black-swan downswing, but the new availability of nest sites resulted in a rapid population upswing \citep{potts1980}. As another example, an interaction between environmental- and predation-mediated population cycles are thought to have caused a downward black-swan event for a water vole (\textit{Arvicola terrestris}) population \citep{saucy1994}. Other black swans were the result of a sequence of extreme climate events on their own. For example, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\textit{Ardea cinerea}) abundance in the United Kingdom \citep{stafford1971}. Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted \citep{stafford1971}. \textit{(TODO: substitute a parasite example.)}

@@

We found strong evidence for black swans (heavy-tailed process noise) in 3--5\% of ecological time series. Black swans are usually downward events and are detected more frequently in longer time series and in populations with a smaller magnitude of process error. However, black swans were not associated with density dependence, population growth rate, or lifespan. In verified cases, black swans were often a result of the interaction between elements of extreme climate, predators and parasite cycles, and strong changes in human pressures. Our empirical results, sensitivity analyses, and simulation tests suggest that estimating the tail shape of process noise is a viable method of detecting black-swan population dynamics and if anything will underestimate the probability of black swans. The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century \citep{easterling2000,ipcc2012}.

@@

There are a number of caveats when considering the generality of our results. The GPDD data represent a taxonomically and geographically biased sample of populations --- the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom. Although we expect that we would find qualitatively similar evidence for black swans in many other large taxonomic or geographic samples of populations, the common forces driving those black swans likely differ (e.g.~severe winters in Table~\ref{tab:sparks}). In addition to a possibly biased sample of populations, some black-swan detections could just be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes. Indeed, three of the populations our method initially identified as heavy tailed turned out to be data-entry errors (Supporting Information). A final caveat is that the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and this likely influences the detection of heavy tails. For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black swans. Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.
